Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-09-13 Code/Manuscript-SMOTE.Rmd
updating

### Import libraries
ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'foreach'))

@@

Cause of death codes are very broad (e.g., "insect" or "disease"). Accuracy of FIA data is commonly assessed using blind checks, where two crews perform independent inventories. Agreement between mortality agent codes recorded in the two inventories is generally >80% [@anderegg_tree_2015]. Active damage is easier to identify, thus the codes for damaging agents are much more specific.
```{r determine presence of BB agents in focal hosts, eval=F}
datz5$WBBB.l <- NA
datz5$WBBB.d <- NA
datz5$WBBB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD==ABLA |datz5$DAMAGE_AGENT_CD1 %in% c(wbbb, sfm) & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB <- rowSums(datz5[,c("WBBB.d", "WBBB.l")], na.rm=T)
datz5$WBBB[is.na(datz5$WBBB)==T]<-0

@@

#### Regional bark beetle pressure
Additionally, given the effects of stand structure and composition on the susceptibility to bark beetle infestation are expected to be greatest in the initial stages of outbreak development [@raffa_cross-scale_2008, @hart_summer_2017], we calculated a proxy of regional bark beetle population density. To do so, we first identified all plots located within 25 km of the focal plot (hereafter *neighboring plots*). We then calculated the regional bark beetle pressure as:

$BP_{regional} = BA_{infested}/BA_{host_{inventoried}} * BA_{host_{total}}$

where $BP_{regional}$ is the regional beetle pressure, $BA_{infested}$ is the total basal area of infested trees within neighboring plots inventoried in the focal or preceding years, $BA_{host_{inventoried}}$ is the total basal area of host trees within neighboring plots inventoried in same year or preceding years, $BA_{host_{total}}$ is the total basal area of host trees for all neighboring plots, $n_{plots}$ is the number of neighboring plots.

```{r pressure, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-df.csv"))
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))

plots <- unique(df[,c("PLT_CN", "INVYR", "LAT", "LON")])
plotspdf <- SpatialPointsDataFrame(plots[,c("LON", "LAT")], data=plots, proj4string=CRS("+init=epsg:4269"))

plotspdf.t <- spTransform(plotspdf,
CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"))

# calculate distances between plots
plot.dists <- rgeos::gDistance(plotspdf.t, byid=T)
colnames(plot.dists) <- plotspdf$PLT_CN
rownames(plot.dists) <- plotspdf$PLT_CN
pressure <- data.frame(PLT_CN=plotspdf$PLT_CN, hostBA.MPB=NA, hostBA.SB=NA, hostBA.WBBB=NA, hostBA.MPB.samp=NA, hostBA.SB.samp=NA, hostBA.WBBB.samp=NA, hostBA.allBBs.samp=NA, nplots=NA, pressureMPB=NA, pressureSB=NA, pressureWBBB=NA, pressureallBBs=NA)

for(j in 1:ncol(plot.dists)){

  # determine plots within 25 km 
  x <- plot.dists[,j] 
  x <- x[-j] # remove plot j 
  x25 <- x[x<=25000]

  # set inventory year for plot j
  invyr <- plots[j, "INVYR"]

  # select plots that were inventoried before plot j
  pre_PLT_CNs <- plots[plots$INVYR<=invyr, "PLT_CN"]

  # select plots within 25 km that were also inventoried before plot j
  x25r <- x25[names(x25) %in% pre_PLT_CNs]

  # calculate area of host basal area within 25 km using only the most recent plot inventories 
  if(length(x25)>=0){
    pressure[j,  "hostBA.MPB"] <- sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.PICO"], na.rm=T) 
    pressure[j,  "hostBA.SB"] <-sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.PIEN"], na.rm=T) 
    pressure[j,  "hostBA.WBBB"] <- sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.ABLA"], na.rm=T) 
    pressure[j,  "hostBA.allBBs"] <- sum(rowSums(df.recent[df.recent$PLT_CN %in% names(x25), c( "BA.PICO", "BA.PIEN", "BA.ABLA")], na.rm=T), na.rm=T) 
  }

  # calculate regional pressure
  if(length(x25r)>0){
    df.x25r <- df[df$PLT_CN %in% names(x25r),]

    # calculate total basal area of PICO, PIEN, and ABLA sampled for all plots within 25 km sampled in the current or previous years 
    pressure[j,  "hostBA.MPB.samp"]  <- sum(df.x25r[, "BA.PICO"], na.rm=T) 
    pressure[j,  "hostBA.SB.samp"] <-sum(df.x25r[, "BA.PIEN"], na.rm=T) 
    pressure[j,  "hostBA.WBBB.samp"] <- sum(df.x25r[, "BA.ABLA"], na.rm=T) 
    pressure[j,  "hostBA.allBBs.samp"]   <- sum(rowSums(df.x25r[, c( "BA.PICO", "BA.PIEN", "BA.ABLA")], na.rm=T), na.rm=T)

    # calculate number of plots sampled for all plots within 25 km sampled in the current or previous years 
    pressure[j,  "nplots"]   <- nrow(df.x25r)

   # calculate total basal area of trees affected by MPB, SB, and WBBB sampled for all plots within 25 km sampled in the current or previous years 
    pressure[j,  "pressureMPB"] <- sum(df.x25r[, "BA.MPB"], na.rm=T)  
    pressure[j,  "pressureSB"] <- sum(df.x25r[, "BA.SB"], na.rm=T)   
    pressure[j,  "pressureWBBB"] <- sum(df.x25r[, "BA.WBBB"], na.rm=T) 
    pressure[j,  "pressureallBBs"] <- sum(df.x25r[, c("BA.MPB", "BA.SB", "BA.WBBB")], na.rm=T) 
  }

}

pressure$Rpressure.MPB <- (pressure$pressureMPB/pressure$hostBA.MPB.samp) * pressure$hostBA.MPB 
pressure$Rpressure.SB <- (pressure$pressureSB/pressure$hostBA.SB.samp) *pressure$hostBA.SB
pressure$Rpressure.WBBB <- (pressure$pressureWBBB/pressure$hostBA.WBBB.samp) *pressure$hostBA.WBBB 
pressure$Rpressure.allBBs <- ( pressure$pressureallBBs/pressure$hostBA.allBBs.samp) * pressure$hostBA.allBBs
pressure[is.na(pressure)==T]<-0

pressure <- pressure[pressure$PLT_CN %in% df.recent$PLT_CN,]
write.csv(pressure, paste0(here("Data/Processed/FIAdat-pressure.csv")),row.names=F)
```

#### Climate data
Finally, because bark beetle developmental rates and overwintering mortality are sensitive to summer and winter temperatures [@miller_cold-hardiness_1987, @bentz_temperature-dependent_1991, @hansen_temperature-based_2001], respectively, we acquired the average maximum temperatures for July (JulTMAX) and average minimum temperatures for December (DecTMIN). Because prior research shows associations between precipitation and outbreak [@hebertson_climate_2008; @chapman_spatiotemporal_2012], we also obtained monthly total precipitation and calculated the total water year (i.e. September to October) precipitation (WYPPT). As indices of host stress due to drought, we used the Standardized Evapotranspiration Precipitation Index (SPEI). SPEI is a multiscalar index that uses precipitation and temperature data in a simple Thornthwaite water balance model [@thornthwaite_approach_1948] to quantify drought [vicente-serrano_multiscalar_2009,]. Specifically we obtained SPEI calculated over the June–July period (JJSPEI) and a 5-yr period, which spanned September to the October five years prior (i.e., five water years; WYSPEI) [@hart_summer_2017]. SPEI data were produced from PRISM data and downloaded from the West Wide Drought Tracker [@abatzoglou_west_2017]. For each plot, we calculated the 1982-2017 climatic normals. For each plot we assigned the climate normal value of the 4 x 4 km pixel underlying each point.

```{r download and process climate data, eval=F}
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
plots <- unique(df.recent[,c("PLT_CN", "LAT", "LON", "INVYR")])
plotspdf <- SpatialPointsDataFrame(plots[,c("LON", "LAT")], data=plots, proj4string=CRS("+init=epsg:4269"))

# PRISM
dir.create(here("Data/Spatial/PRISM/"), showWarnings = FALSE)

## Normals
dir.create(here("Data/Spatial/PRISM/Normals/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/Normals/"))
get_prism_normals("ppt", resolution="4km", annual = TRUE)
get_prism_normals("tmax", resolution="4km", mon=7, annual = FALSE)
get_prism_normals("tmin", resolution="4km", mon=12, annual = FALSE)

normals <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))
normals.df <- raster::extract(normals, plotspdf, method="simple", df=T)
normals.df$PLT_CN <- plots$PLT_CN
colnames(normals.df) <- c("nPPT12", "nTMAX1.7", "nTMIN1.12", "PLT_CN")
write.csv(normals.df, here("Data/Processed/Normals.csv"), row.names = F)

## Jul MAX Temperature
dir.create(here("Data/Spatial/PRISM/TMAX1_7/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/TMAX1_7/"))
tmax <- get_prism_monthlys(type="tmax", mon=7, years=1981:2020)
TMAX1.7 <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))

options(prism.path=here("Data/Spatial/PRISM/Normals/"))
nTMAX1.7 <- raster(pd_to_file(prism_archive_ls()[2]))
TMAX1.7.anom <- (TMAX1.7+100) - (nTMAX1.7+100)

TMAX1.7.df <- raster::extract(TMAX1.7, plotspdf, method="simple", df=T)
TMAX1.7.df$PLT_CN <- plots$PLT_CN
write.csv(TMAX1.7.df, here("Data/Processed/TMAX1-7.csv"), row.names = F)

TMAX1.7.anom.df <- raster::extract(TMAX1.7.anom, plotspdf, method="simple", df=T)
TMAX1.7.anom.df$PLT_CN <- plots$PLT_CN
write.csv(TMAX1.7.anom.df, here("Data/Processed/TMAX1-7anom.csv"), row.names = F)

## Dec MIN Temperature
dir.create(here("Data/Spatial/PRISM/TMIN1_12/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/TMIN1_12/"))
TMIN12 <- get_prism_monthlys(type="tmin", mon=12, years=1981:2020)
TMIN1.12 <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))

options(prism.path=here("Data/Spatial/PRISM/Normals/"))
nTMIN1.12 <- raster(pd_to_file(prism_archive_ls()[3]))
TMIN1.12.anom <- (TMIN1.12+100) - (nTMIN1.12+100)

tmin1.12.df <- raster::extract(TMIN1.12, plotspdf, method="simple", df=T)
tmin1.12.df$PLT_CN <- plots$PLT_CN
write.csv(tmin1.12.df, here("Data/Processed/TMIN1-12.csv"), row.names = F)

tmin1.12.anom.df <- raster::extract(TMIN1.12.anom, plotspdf, method="simple", df=T)
tmin1.12.anom.df$PLT_CN <- plots$PLT_CN
write.csv(tmin1.12.anom.df, here("Data/Processed/TMIN1-12anom.csv"), row.names = F)

## Annual PPT
dir.create(here("Data/Spatial/PRISM/PPT12/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/PPT12/"))
get_prism_annual(type="ppt", years=1981:2020)
ppt <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))

options(prism.path=here("Data/Spatial/PRISM/Normals/"))
nppt <- raster(pd_to_file(prism_archive_ls()[1]))
pptanom <- ppt - nppt

ppt.df <- raster::extract(ppt, plotspdf, method="simple", df=T)
ppt.df$PLT_CN <- plots$PLT_CN
write.csv(ppt.df, here("Data/Processed/PPT12.csv"), row.names = F)

pptanom.df <- raster::extract(pptanom, plotspdf, method="simple", df=T)
pptanom.df$PLT_CN <- plots$PLT_CN
write.csv(pptanom.df, here("Data/Processed/PPT12anom.csv"), row.names = F)

## JUN-JUL SPEI
dir.create(here("Data/Spatial/PRISM/SPEI2_7"), showWarnings = FALSE)
temp <- here("/Data/Spatial/PRISM/spei2_7_PRISM.nc")
download.file("http://www.wrcc.dri.edu/wwdt/data/PRISM/spei2/spei2_7_PRISM.nc", temp)

spei2 <- brick(here("Data/Spatial/PRISM/SPEI2_7/spei2_7_PRISM.nc"))
spei2.df <-raster:::extract(spei2, plotspdf, method="simple", df=T)[,-1]
spei2.df$PLT_CN <- plots$PLT_CN
write.csv(spei2.df,paste0(here("Data/Processed"),"SPEI2.csv"), row.names = F)

## 5 WATER YEAR SPEI
dir.create(here("Data/Spatial/PRISM/SPEI60_9"), showWarnings = FALSE)
temp <- here("Data/Spatial/PRISM/SPEI60_9/spei60_9_PRISM.nc")
download.file("http://www.wrcc.dri.edu/wwdt/data/PRISM/spei60/spei60_9_PRISM.nc", temp)

spei60 <- brick( here("Data/Spatial/PRISM/SPEI60_9/spei60_9_PRISM.nc"))
spei60.df <-raster:::extract(spei60, plotspdf, method="simple", df=T)[,-1]
spei60.df$PLT_CN <- plots$PLT_CN
write.csv(spei60.df,paste0(here("Data/Processed/"),"SPEI60.csv"), row.names = F)

spei60.df <- read.csv(here("Data/Processed/SPEI60.csv"))
spei2.df <- read.csv(here("Data/Processed/SPEI2.csv"))

ppt.df <- read.csv( here("Data/Processed/PPT12.csv"))
pptanom.df <- read.csv( here("Data/Processed/PPT12anom.csv"))

tmin1.12.df <- read.csv(here("Data/Processed/TMIN1-12.csv"))
tmin1.12.anom.df <- read.csv(here("Data/Processed/TMIN1-12anom.csv"))

tmax1.7.df <- read.csv(here("Data/Processed/TMAX1-7.csv"))
tmax1.7.anom.df <- read.csv(here("Data/Processed/TMAX1-7anom.csv"))

climatedata <- plots  
for(j in 1:nrow(climatedata)){
  INVYR <- climatedata[j, "INVYR"]
  end.yr <- INVYR 
  start.yr <- INVYR-9

  spei2.ts <- ts(t(spei2.df[j,-1*ncol(spei2.df)]),start=1895, freq=1)
  spei2.tsw <- window(spei2.ts, end=end.yr, start=start.yr)
  climatedata[j, "JJSPEI"] <- mean(spei2.tsw , na.rm=T)

  spei60.ts <- ts(t(spei60.df[j,-1*ncol(spei60.df)]),start=1895, freq=1)
  spei60.tsw <- window(spei60.ts, end=end.yr, start=start.yr)
  climatedata[j, "WYSPEI"] <- mean(spei60.tsw , na.rm=T)

 tmin1.12.ts <- ts(t(tmin1.12.df[j,-1*ncol(tmin1.12.df )]),start=1981, freq=1)
  tmin1.12.tsw <- window(tmin1.12.ts, end=end.yr, start=start.yr)
  climatedata[j, "DecTMIN"] <- mean(tmin1.12.tsw[,1], na.rm=T)

  tmin1.12a.ts <- ts(t(tmin1.12.anom.df[j,-1*ncol(tmin1.12.anom.df )]),start=1981, freq=1)
  tmin1.12a.tsw <- window(tmin1.12a.ts, end=end.yr, start=start.yr)
  climatedata[j, "DecTMINa"] <- mean(tmin1.12a.tsw[,1], na.rm=T)

   tmax1.7.ts <- ts(t(tmax1.7.df[j,-1*ncol(tmax1.7.df )]),start=1981, freq=1)
  tmax1.7.tsw <- window(tmax1.7.ts, end=end.yr, start=start.yr)
  climatedata[j, "JulTMAX"] <- mean(tmax1.7.tsw[,1], na.rm=T)

tmax1.7a.ts <- ts(t(tmax1.7.anom.df[j,-1*ncol(tmax1.7.anom.df )]),start=1981, freq=1)

  tmax1.7a.tsw <- window(tmax1.7a.ts, end=end.yr, start=start.yr)
  climatedata[j, "JulTMAXa"] <- mean(tmax1.7a.tsw[,1], na.rm=T)

  ppt.ts <- ts(t(ppt.df [j,-1*ncol(ppt.df )]),start=1981, freq=1)
  ppt.tsw <- window( ppt.ts, end=end.yr, start=start.yr)
  climatedata[j, "PPT"] <- mean(ppt.tsw[,1], na.rm=T)

  ppta.ts <- ts(t(pptanom.df [j,-1*ncol(pptanom.df )]),start=1981, freq=1)
 ppta.tsw <- window(ppta.ts, end=end.yr, start=start.yr)
  climatedata[j, "PPTa"] <- mean(ppta.tsw[,1], na.rm=T)
}

write.csv(climatedata, here("Data/Processed/climatedata.csv"), row.names = F)
```

```{r merge data, eval=F}
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
pressure <- read.csv(here("Data/Processed/FIAdat-pressure.csv"))
climatedata <- read.csv(here("Data/Processed/climatedata.csv"))

df <- merge(df.recent, pressure, by="PLT_CN", all=T)
df <- merge(df, climatedata[, -2:-4], by="PLT_CN", all=T)

write.csv(df, here("Data/Processed/FIAdat-df-merge.csv"), row.names=F)
```

@@

  df.p$BAD.host <- df.p$BA.host/df.p$BA.total
  df.p$TPHD.host <- df.p$TPH.host/df.p$TPH.total
  df.p$pressure.BB <- df.p[,paste0("pressure", agent)]

@@

# Descriptive analyses
```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), ordered=T)
df$hostsFPlong <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

@@

### Model occurrence

```{r select plots to control for sampling effort, eval=T}
#df.sub <- df[df$nplots>5, ]
df.sub <- df
## remove plots with other hosts
#df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
#nrow(datBB.c)
df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]

# select plots where focal agent is all

df.sub <- df.sub[df.sub$focalagent=="Any", ]
```

```{r select plots with large & abundant hosts, eval=T}
threshold <- 0.10
PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, threshold)
PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, threshold)
ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, threshold)

PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, threshold)
PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, threshold)
ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, threshold)

df.sub$hosts2.PICO <- "absent"
df.sub[df.sub$QMD.PICO >PICO.QMD.threshold & df.sub$BA.PICO >PICO.BA.threshold, "hosts2.PICO"] <- "PICO"
df.sub$hosts2.PIEN <- "absent"
df.sub[df.sub$QMD.PIEN >PIEN.QMD.threshold & df.sub$BA.PIEN >PIEN.BA.threshold, "hosts2.PIEN"] <- "PIEN"
df.sub$hosts2.ABLA<- "absent"
df.sub[df.sub$QMD.ABLA >ABLA.QMD.threshold & df.sub$BA.ABLA >ABLA.BA.threshold, "hosts2.ABLA"] <- "ABLA"

df.sub$hosts2<- NA
df.sub$hosts2<- do.call(paste, c(df.sub[c("hosts2.PICO", "hosts2.PIEN", "hosts2.ABLA")],sep="-"))
df.sub <- df.sub[!df.sub$hosts2 =='absent-absent-absent',]
df.sub$hosts2 <- factor(df.sub$hosts2, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA'), ordered=TRUE)
df.sub$nhosts2 <- as.numeric(factor(df.sub$hosts2, levels=c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA'), labels=c(1, 1, 1, 2,2,2, 3)))

### Number of hosts

#### Any hosts
affectednhosts2 <- df.sub %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts2 <- affectednhosts2[affectednhosts2$affected=="affected",]
colnames(affectednhosts2) <- c("nhosts", "agent", "n", "freq")
affectednhosts2$agent <- "Any"

#### MPB
MPBnhosts2 <- df.sub %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
MPBnhosts2 <- MPBnhosts2[MPBnhosts2$agentMPB=="MPB",]
colnames(MPBnhosts2) <- c("nhosts", "agent", "n", "freq")

#### SB
SBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentSB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
SBnhosts2 <- SBnhosts2[SBnhosts2$agentSB=="SB",]
colnames(SBnhosts2) <- c("nhosts", "agent", "n", "freq")

#### WBBB
WBBBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentWBBB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
WBBBnhosts2 <- WBBBnhosts2[WBBBnhosts2$agentWBBB=="WBBB",]
colnames(WBBBnhosts2) <- c("nhosts", "agent", "n", "freq")

#### Merge
affectednhosts2 <- rbind(as.data.frame(affectednhosts2),as.data.frame(MPBnhosts2))
affectednhosts2 <- rbind(affectednhosts2, as.data.frame(SBnhosts2))
affectednhosts2 <- rbind(affectednhosts2, as.data.frame(WBBBnhosts2))

### Host identity
#### Any hosts
affectedhosts2 <- df.sub %>% group_by(hosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhosts2 <- affectedhosts2[affectedhosts2$affected=="affected",]
colnames(affectedhosts2) <- c("hosts", "agent", "n", "freq")
affectedhosts2$agent <- "Any"

#### MPB
MPBhosts2 <- df.sub %>% group_by(hosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
MPBhosts2 <- MPBhosts2[MPBhosts2$agentMPB=="MPB",]
colnames(MPBhosts2) <- c("hosts", "agent", "n", "freq")

#### SB
SBhosts2 <- df.sub  %>% group_by(hosts2, agentSB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
SBhosts2 <- SBhosts2[SBhosts2$agentSB=="SB",]
colnames(SBhosts2) <- c("hosts", "agent", "n", "freq")

#### WBBB
WBBBhosts2 <-df.sub  %>% group_by(hosts2, agentWBBB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
WBBBhosts2 <- WBBBhosts2[WBBBhosts2$agentWBBB=="WBBB",]
colnames(WBBBhosts2) <- c("hosts", "agent", "n", "freq")

#### Merge
affectedhosts2 <- rbind(as.data.frame(affectedhosts2),as.data.frame(MPBhosts2))
affectedhosts2 <- rbind(affectedhosts2, as.data.frame(SBhosts2))
affectedhosts2 <- rbind(affectedhosts2, as.data.frame(WBBBhosts2))
affectedhosts2$hosts <- factor(affectedhosts2$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

### test for significant differences
affectedhosts3 <- as.data.frame(table(df.sub$hosts2))
colnames(affectedhosts3) <- c("hosts", "N")

affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="Any", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any")
affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="MPB", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB")
affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="SB", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB", "SB")
affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="WBBB", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB", "SB", "WBBB")

affectedhosts3$label <- paste0(do.call(paste, c(affectedhosts3[c("hosts", "n")],sep="\n(n=")), ")")

affectedhosts2$p.value <- NA
affectedhosts2$statistic <- NA
for(j in unique(affectedhosts2$hosts)){
  affectedhosts2j <- affectedhosts2[affectedhosts2$hosts==j, ]
  affectedhosts3j <- affectedhosts3[affectedhosts3$hosts==j,]
  for(k in unique(affectedhosts2j$agent)[!unique(affectedhosts2j$agent)=="Any"]){

    test <- prop.test(x=do.call(c,affectedhosts3j[, c("Any", k)]), n=rep(affectedhosts3j[, "n"],2), alternative="greater")
  affectedhosts2[affectedhosts2$hosts==j & affectedhosts2$agent==k, "p.value"] <- round(test$p.value,3)
    affectedhosts2[affectedhosts2$hosts==j & affectedhosts2$agent==k, "statistic"] <- test$statistic
  }
}
affectedhosts2$symbol <- ifelse( affectedhosts2$p.value<0.05, "*", "")
affectedhosts2$agent <- factor(affectedhosts2$agent, levels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

affectedhosts2 <- affectedhosts2[affectedhosts2$hosts %in% c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ]
affectedhosts2$hosts <- factor(affectedhosts2$hosts, levels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=affectedhosts3$label[c(4,3,7,5)], ordered=T)

Fig.file <- here("Results/Figures/Figure-OccurenceGivenSuitableStands.jpg")
jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
ggplot(affectedhosts2, aes(x=agent, y=freq, fill=agent))+geom_bar(stat="identity")+
geom_text(aes(label = symbol), vjust = 0.2)+facet_wrap(.~hosts, scales="free", nrow=1)+ylab("proportion affected")+xlab("bark beetle species")+scale_fill_manual(values=c(rev(p2col)[2:4], grey(0.25)))+theme(legend.position="none")
whatever <- dev.off()

```{r, eval=T}
datBB.c <- df.sub[df.sub$focalhost=="Any",]
splitit <- initial_split(df.sub, prop=0.5)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, v=10)
```

```{r fit MPB RF model, eval=T}
response <- "agentMPB"
#vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO", "pressureMPB", "hostBA.MPB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")

form <- paste(response, "~", paste(vars, collapse = " + "))

>>>>

### Import libraries
ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach'))

@@

Cause of death codes are very broad (e.g., "insect" or "disease"). Accuracy of FIA data is commonly assessed using blind checks, where two crews perform independent inventories. Agreement between mortality agent codes recorded in the two inventories is generally >80% [@anderegg_tree_2015]. Active damage is easier to identify, thus the codes for damaging agents are much more specific.

```{r determine presence of BB agents in focal hosts, eval=F}
datz5$WBBB.l <- NA
datz5$WBBB.d <- NA
datz5$WBBB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD==ABLA |datz5$DAMAGE_AGENT_CD1 %in% c(wbbb, sfm) & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB <- rowSums(datz5[,c("WBBB.d", "WBBB.l")], na.rm=T)
datz5$WBBB[is.na(datz5$WBBB)==T]<-0

@@

```{r merge data, eval=F}
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
write.csv(df, here("Data/Processed/FIAdat-df-merge.csv"), row.names=F)
```

@@

  df.p$BAD.host <- df.p$BA.host/df.p$BA.total
  df.p$TPHD.host <- df.p$TPH.host/df.p$TPH.total
  #df.p$pressure.BB <- df.p[,paste0("pressure", agent)]

@@

### Descriptive analyses
```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), ordered=T)
df$hostsFPlong <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

@@

```{r}
Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
df$hostsFPshort<- df$hostsFPlong
df$hostsFPshort <- factor(df$hostsFPlong, levels=levels(df$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
df$agentsFplongrev <- df$agentsFplong
df$agentsFplongrev <- factor(df$agentsFplongrev, levels=rev(levels( df$agentsFplong)))
jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
ggplot(df[df$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
whatever <- dev.off()
```

### Given suitable stand conditions, is co-occurrence common?

Construct random forest models using synthetic minority oversampling technique (SMOTE)

#### MPB

```{r fit MPB RF model, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)

# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
